The Design, Semantics, and Implementation of \CEU: a Synchronous Reactive Language based on Esterel

### 2015-10-08 paper.tex
Revisao ate of final Mais texto do Adriano

The core of our semantics is a relation that, given a sequence number $n$ 
identifying the current reaction chain, maps a program $p$ and a stack of 
events $S$ in a single step to a modified program and stack:
%
\begin{align*}
\LL S, p \RR &\ST
\LL S', p' \RR
    & \textbf{(relation-inner)}
\end{align*}
%
where
%
\begin{align*}
S, S' &\in id^*
    &&(sequence~of~event~identifiers: [id_{top}, ..., id_{bottom}]) \\
p, p' &\in P
    && (program~as~described~in~Figure~\ref{lst.formal.syntax}) \\
n     &\in \mathds{N}
    && (unique~identifier~for~the~reaction~chain)
\end{align*}
%
At the beginning of a reaction chain, the stack is initialized with the 
occurring external event $ext$ ($S=[ext]$), but $emit$ expressions can push new 
events on top of it (we discuss how they are popped further).
The ever-increasing sequence number $n$ prevents that awaiting expressions 
awake in the same reaction they are reached (the \emph{delayed awaits} as 
explained in Section~\ref{sec.ceu.ints}).

@@

Proceeding to compound expressions, the rules for conditionals and sequences 
are straightforward:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \frac
    { \DS val(id,n) \neq 0 }
%   -----------------------------------------------------------
    { \DS \LL S, (if~mem(id)~then~p~else~q) \RR \ST
          \LL S, p \RR }
    & \textbf{(if-true)}       \\
%%%
& \frac
    { \DS val(id,n) = 0 }
%   -----------------------------------------------------------
    { \DS \LL S, (if~mem(id)~then~p~else~q) \RR \ST
          \LL S, q \RR }
    & \textbf{(if-false)}       \\
%%%
& \frac
    {\DS \LL S,p \RR \ST \LL S',p' \RR }
%   -----------------------------------------------------------
    {\DS \LL S, (p~;~q) \RR \ST \LL S', (p'~;~q) \RR }
    & \textbf{(seq-adv)}      \\
%%%
& \LL S, (mem(id)~;~q) \RR \ST  \LL S, q \RR
    & \textbf{(seq-nop)}      \\
%%%
& \LL S, (break~;~q) \RR \ST \LL S, break \RR
    & \textbf{(seq-brk)}
\end{eqnarray*}
}
%
Given that our semantics focuses on control, rules \textbf{if-true} and 
\textbf{if-false} are the only to query $mem$ expressions.
%
The ``magical'' function $val$ receives a memory identifier and the current 
reaction sequence number, returning the current memory value.
%
Although the value here is arbitrary, it is unique in a reaction chain, because 
a given expression can execute only once within it (remember that $loops$ must 
contain $awaits$ which, from rule \textbf{await}, cannot awake in the same 
reaction they are reached).
%For all other rules, we omit these values (e.g., \textbf{seq-nop}).

@@

For a parallel $and$, if one of the sides terminates, the composition is simply 
substituted by the other side (rules \textbf{and-nop1} and \textbf{and-nop2}, 
as follows).
%
For a parallel $or$, if one of the sides terminates, the whole composition 
terminates, also applying the $clear$ function to properly finalize the aborted 
side (rules \textbf{or-nop1} and \textbf{or-nop2}):
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \LL S, (mem(id)~and~q) \RR \ST \LL S, q \RR
    & \textbf{(and-nop1)}   \\
%%%
& \LL S, (p~and~mem(id)) \RR \ST \LL S, p \RR
    & \textbf{(and-nop2)}   \\
%%%
& \LL S, (mem(id)~or~q) \RR \ST \LL S, clear(q) \RR
    & \textbf{(or-nop1)}   \\
%%%
& \frac
    {\DS isBlocked(n,S,p) }
%   -----------------------------------------------------------
    {\DS \LL S, (p~or~mem(id)) \RR \ST \LL S, clear(p) \RR }
    & \textbf{(or-nop2)}   %\\
\end{eqnarray*}
}
%
The $clear$ function, defined in Figure~\ref{fig.formal.clear}, concatenates 
all active $fin$ bodies of the side being aborted, so that they execute before 
the composition rejoins.
Note that there are no transition rules for $fin$ expressions.
This is because once reached, a $fin$ expression halts and will only execute 
when it is aborted by a trail in parallel and is expanded by the $clear$ 
function.
In Section~\ref{sec.formal.fins}, we show how to map a finalization block in 
the concrete language to a $fin$ in the formal semantics.
%
Note that there is a syntactic restriction that a $fin$ body can only contain 
$mem$ expressions, i.e., they are guaranteed to execute entirely within a 
reaction chain.

@@

Finally, a $break$ in one of the sides in parallel escapes the closest 
enclosing $loop$, properly aborting the other side by applying the $clear$ 
function:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \LL S, (break~and~q) \RR \ST \LL S, (clear(q)~;~break) \RR
    & \textbf{(and-brk1)}   \\
%%%
& \frac
    {\DS isBlocked(n,S,p) }
%   -----------------------------------------------------------
    {\DS \LL S, (p~and~break) \RR \ST \LL S, (clear(p)~;~break) \RR }
    & \textbf{(and-brk2)}   \\
%%%
& \LL S, (break~or~q) \ST \LL S, (clear(q)~;~break) \RR
    & \textbf{(or-brk1)}   \\
%%%
& \frac
    {\DS isBlocked(n,S,p) }
%   -----------------------------------------------------------
    {\DS \LL S, (p~or~break) \RR \ST \LL S, (clear(p)~;~break) \RR }
    & \textbf{(or-brk2)}   %\\
\end{eqnarray*}
}
%
A reaction chain eventually blocks in $awaiting$ and $emitting$ expressions in 
parallel trails.
%
If all trails hangs only in $awaiting$ expressions, it means that the program 
cannot advance in the current reaction chain.
%
However, $emitting$ expressions are pending in lower stack indexes and should 
eventualy resume in the ongoing reaction (see rule \textbf{pop}).
%
Therefore, we define another relation that behaves as \textbf{relation-inner} 
(presented above) if the program can advance, and, otherwise, pops the stack:
%
$$
\frac
    { \DS \LL S,p \RR \ST                   \LL S',p' \RR }
%   -----------------------------------------------------------
    {     \LL S,p \RR \xRightarrow[~~n~~]{} \LL S',p' \RR }
%
\5\5\5
%
\frac
    { \DS isBlocked(n,\1s:S,\1p) }
%   -----------------------------------------------------------
    { \LL s:S,p \RR \xRightarrow[~~n~~]{} \LL S,p \RR }
\5\5\5\textbf{(relation-outer)}
$$
%
To describe a \emph{reaction chain} in \CEU, i.e., how a program behaves in 
reaction to a single external event, we use the reflexive transitive closure of 
\textbf{relation-outer}:
%
$$
    \LL S,p \RR \xRightarrow[~~n~~]{*} \LL S',p' \RR
$$
%
Finally, to describe the complete execution of a program, we trigger multiple 
``invocations'' of reaction chains in sequence:
%
\begin{align*}
\LL [e1], p \RR
    & \xRightarrow[~~1~~]{*}
\LL [  ], p' \RR
\\
\LL [e2], p' \RR
    & \xRightarrow[~~2~~]{*}
\LL [  ], p'' \RR
\\
\LL [e3], p'' \RR
    & \xRightarrow[~~3~~]{*}
\LL [  ], p''' \RR
\\
& ...
\end{align*}
%
Each invocation starts with an external event at the top of the stack and 
finishes with a modified program and an empty stack.
After each invocation, we increment the sequence number.

@@

The concrete \code{await} and \code{emit} primitives support communication of 
values between them.
In the two-step translation of Figure~\ref{lst.map.emit.await}, we start with 
the concrete program in \CEU, which communicates the value $1$ between the 
\code{emit} and \code{await} in parallel (left-most code).
In the intermediate translation, we include the shared variable \code{e\_} to 
hold the value being communicated between the two trails in order to simplify 
the \code{emit}.
Finally, we convert the program into the equivalent in the abstract syntax, 
translating side-effect statements into $mem$ expressions.
External events require a similar translation, i.e., each external event has a 
corresponding variable that is explicitly set by the environment before each 
reaction chain.

@@

In the final generated code in \C, each trail segment label representing an 
entry point becomes a \emph{switch case} with the associated code to execute.
%
Figure~\ref{lst.impl.trails} illustrates the generation process.
For the program in the left of the figure, the compiler extracts the entry 
points and associated trails, e.g., the label \code{Awake\_e} will execute on 
\code{TRAIL-0} (ln. 7).
%
For each statement that pauses (\code{emit} and \code{await}), resumes 
(\code{par/and}, \code{par/or}, and \code{finalize}), or aborts (\code{par/or} 
and \code{break}), the compiler splits the trail into segments with associated 
entry points.
%
The entry points translate to an \code{enum} in the generated code (ln. 1--10, 
in the middle of the figure).
The state of trails translate to a vector of type \code{trail\_t} with the 
maximum number of simultaneous trails (ln. 12--15).
On initialization, \code{TRAIL-0} is set to execute the \code{Main} entry point 
(ln. 13), while all others are set to \code{INACTIVE} (in the example, only 
one, in ln. 14).

@@

The scheduler executes in two passes:
%
In the \emph{broadcast} pass, the scheduler sets all trails that are waiting 
for the current event to \code{STACKED} in the current stack level.
%
In the \emph{dispatch} pass, the scheduler executes each trail that is 
\code{STACKED} to run in the current level, setting it immediately to 
\code{INACTIVE}.

During the dispatch pass, if a trail executes and emits an internal event, the 
scheduler increments the stack level and re-executes the two passes.
After all trails are properly dispatched, the scheduler decrements the stack 
level and resumes the previous execution.
For the first reaction chain, the scheduler starts from the \emph{dispatch} 
pass, given that the \code{Main} label is the only one that can be active at 
the stack level 0 (ln. 13, in the middle of Figure~\ref{lst.impl.trails}).

The code the right of the Figure~\ref{lst.impl.trails} dispatches a trail
according to the current label to execute.
%
For the first reaction, it executes the \code{Main} label in \code{TRAIL-0}.
%
When the \code{Main} label reaches the \code{par/and}, it stacks \code{TRAIL-1} 
(ln. 4--7) and proceeds to the code in \code{TRAIL-0} (ln. 9--14), respecting 
the deterministic execution order.
%
The code sets the running \code{TRAIL-0} to await \code{EVT\_e} on label 
\code{Awake\_e}, and then halts with a \code{break}.
%
The next iteration of \code{dispatch} takes \code{TRAIL-1} and executes its 
registered label \code{And\_sub\_2} (ln. 16--21), which sets \code{TRAIL-1} to 
await \code{EVT\_A} and also halts.

@@

\begin{description}
\item[{\textbf\code{ceu\_go\_init()}}:] initializes the program (e.g. trails) 
and executes the ``boot'' reaction (i.e., the \code{Main} label).

\item[{\textbf\code{ceu\_go\_event(id,param)}}:] executes the reaction for the 
received event id and associated parameter.

@@

The bindings for the specific platforms are responsible for calling the 
functions in the API in the order that better suit their requirements.
As an example, it is possible to set different priorities for events that occur 
concurrently (i.e. while a reaction chain is running).
However, a binding must never interleave or run multiple functions in parallel.
This would break the \CEU sequential/discrete semantics of time.
%, as discussed in Section~\ref{sec:ceu}.

\begin{figure}
\begin{lstlisting}[numbers=left,xleftmargin=2em]
implementation
{
    #include "ceu.h"
    #include "ceu.c"

@@

As an example, Figure~\ref{lst.impl.tinyos} shows our binding for 
\emph{TinyOS}~\cite{wsn.tos}, which maps callbacks to input events in \CEU.
%
The file \code{ceu.h} (ln. 3) contains all definitions for the compiled \CEU 
program, which are further queried through \code{\#ifdef}'s.
The file \code{ceu.c} (ln. 4) contains the main loop of \CEU pointing to the 
labels defined in the program.
The callback \code{Boot.booted} (ln. 6--11) is called by TinyOS on startup, so 
we initialize \CEU inside it (ln. 7).
If the \CEU program uses timers, we also start a periodic timer (ln. 8--10) 
that triggers callback \code{Timer.fired} (ln. 13--17) every 10 milliseconds 
and advances the wall-clock time of \CEU (ln. 15)%
\footnote{We also offer a mechanism to start the underlying timer on demand to
avoid the ``battery unfriendly'' 10ms polling.}.
The remaining lines map pre-defined TinyOS events that can be used in \CEU 
programs, such as the light sensor (ln. 19--23) and the radio transceiver (ln.  
25--36).

@@

@@

\begin{comment}
Instead of generating code in C, Terra generates the vmt assembly code using 
the original Céu labels for trails entry points.
In the final phase for code generation, Terra compiler generates the vmt 
bytecode file using the program address in place of Céu labels.
The use of address allows the Céu scheduler to execute directly a trail without 
a index table to replace the switch-case.
Related to memory layout and trail allocation, Terra uses the same data 
organization approach for all Céu controls and variables.
All Céu data and control structures for a specific program are determined at 
compile time.
Terra arranges all these structures in the beginning of virtual machine memory 
space.
The original Céu run time control functions are maintained with a small 
modification where accesses to control structures have an off-set relative to 
the begin of the virtual machine memory space.
\end{comment}

@@

\CEU has a strong influence from Esterel and embraces the disciplined 
synchronous-reactive model and support for lexical composition of lines of 
execution.
%
However, there are fundamental semantic differences that prevents the design of 
\CEU as pure extensions to Esterel (i.e., on top of its formal semantics).
%
In particular, Esterel has a notion of time similar to that of digital circuits 
in which multiple signals can be active at a clock tick.
In fact, Esterel is also used in hardware design.
%
In \CEU, instead of clock ticks, the occurrence of a single external event that 
defines a time unit.
%
\CEU also distinguishes external events from stack-based internal events, which 
provide a limited form of coroutines supporting reactive statements (e.g., 
\code{await} and \code{par/or}).

@@

A number of synchronous languages have been designed to interoperate with C, 
such as \emph{Reactive C}~\cite{rp.rc}, Protothreads~\cite{wsn.protothreads}, 
\emph{PRET-C}~\cite{rp.pretc} and \emph{SC}~\cite{rp.synchc}.
%
They offer Esterel-like parallel compositions with communication via shared 
variables, relying on deterministic scheduling to preserve determinism.
%
However, it is the responsibility of the programmer to specify the execution 
order for threads, based on either explicit priorities, or source code lexical 
order (like \CEU).
%
These languages have a tick-based notion of time similar to Esterel, which 
prevents the event-based temporal analysis of \CEU.

@@

@@

\CEU is a concurrency-safe language, employing a static analysis that encompass 
all control constructs and ensures that the high degree of concurrency in 
embedded systems does not pose safety threats to applications.
%
As a summary, the following safety properties hold for all programs that 
successfully compile in \CEU:
time and memory-bounded reactions to the environment (except for external 
system calls),
no race conditions in shared memory,
reliable abortion for activities handling resources,
and automatic synchronization for timers.
%
These properties are usually desirable in applications and are guaranteed as 
preconditions in \CEU by design.

>>>>

The core of our semantics is a relation that, given a sequence number $n$ 
identifying the current reaction, maps a program $p$ and a stack of events $S$ 
in a single step to a modified program and stack:
%
\begin{align*}
\LL S, p \RR &\ST
\LL S', p' \RR
    & \textbf{(relation-inner)}
\end{align*}
%
where
%
\begin{align*}
S, S' &\in id^*
    &&(sequence~of~event~identifiers: [id_{top}, ..., id_{bottom}]) \\
p, p' &\in P
    && (program~as~described~in~Figure~\ref{lst.formal.syntax}) \\
n     &\in \mathds{N}
    && (unique~identifier~for~the~reaction)
\end{align*}
%
At the beginning of a reaction, the stack is initialized with the occurring 
external event $ext$ ($S=[ext]$), but $emit$ expressions can push new events on 
top of it (we discuss how they are popped further).
The ever-increasing sequence number $n$ prevents that awaiting expressions 
awake in the same reaction they are reached (the \emph{delayed awaits} as 
explained in Section~\ref{sec.ceu.ints}).

@@

Proceeding to compound expressions, the rules for conditionals and sequences 
are straightforward:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \frac
    { \DS val(id,n) \neq 0 }
%   -----------------------------------------------------------
    { \DS \LL S, (if~mem(id)~then~p~else~q) \RR \ST
          \LL S, p \RR }
    & \textbf{(if-true)}       \\
%%%
& \frac
    { \DS val(id,n) = 0 }
%   -----------------------------------------------------------
    { \DS \LL S, (if~mem(id)~then~p~else~q) \RR \ST
          \LL S, q \RR }
    & \textbf{(if-false)}       \\
%%%
& \frac
    {\DS \LL S,p \RR \ST \LL S',p' \RR }
%   -----------------------------------------------------------
    {\DS \LL S, (p~;~q) \RR \ST \LL S', (p'~;~q) \RR }
    & \textbf{(seq-adv)}      \\
%%%
& \LL S, (mem(id)~;~q) \RR \ST  \LL S, q \RR
    & \textbf{(seq-nop)}      \\
%%%
& \LL S, (break~;~q) \RR \ST \LL S, break \RR
    & \textbf{(seq-brk)}
\end{eqnarray*}
}
%
Given that our semantics focuses on control, rules \textbf{if-true} and 
\textbf{if-false} are the only to query $mem$ expressions.
%
The ``magical'' function $val$ receives a memory identifier and the current 
reaction sequence number, returning the current memory value.
%
Although the value here is arbitrary, it is unique in a reaction, because a 
given expression can execute only once within it (remember that $loops$ must 
contain $awaits$ which, from rule \textbf{await}, cannot awake in the same 
reaction they are reached).
%For all other rules, we omit these values (e.g., \textbf{seq-nop}).

@@

For a parallel $and$, if one of the sides terminates, the composition is simply 
substituted by the other side (rules \textbf{and-nop1} and \textbf{and-nop2}, 
as follows).
%
For a parallel $or$, if one of the sides terminates, the whole composition 
terminates, also applying the $clear$ function to properly finalize the aborted 
side (rules \textbf{or-nop1} and \textbf{or-nop2}):
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \LL S, (mem(id)~and~q) \RR \ST \LL S, q \RR
    & \textbf{(and-nop1)}   \\
%%%
& \LL S, (p~and~mem(id)) \RR \ST \LL S, p \RR
    & \textbf{(and-nop2)}   \\
%%%
& \LL S, (mem(id)~or~q) \RR \ST \LL S, clear(q) \RR
    & \textbf{(or-nop1)}   \\
%%%
& \frac
    {\DS isBlocked(n,S,p) }
%   -----------------------------------------------------------
    {\DS \LL S, (p~or~mem(id)) \RR \ST \LL S, clear(p) \RR }
    & \textbf{(or-nop2)}   %\\
\end{eqnarray*}
}
%
The $clear$ function, defined in Figure~\ref{fig.formal.clear}, concatenates 
all active $fin$ bodies of the side being aborted, so that they execute before 
the composition rejoins.
Note that there are no transition rules for $fin$ expressions.
This is because once reached, a $fin$ expression halts and will only execute 
when it is aborted by a trail in parallel and is expanded by the $clear$ 
function.
In Section~\ref{sec.formal.fins}, we show how to map a finalization block in 
the concrete language to a $fin$ in the formal semantics.
%
Note that there is a syntactic restriction that a $fin$ body can only contain 
$mem$ expressions, i.e., they are guaranteed to execute entirely within a 
reaction.

@@

Finally, a $break$ in one of the sides in parallel escapes the closest 
enclosing $loop$, properly aborting the other side by applying the $clear$ 
function:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \LL S, (break~and~q) \RR \ST \LL S, (clear(q)~;~break) \RR
    & \textbf{(and-brk1)}   \\
%%%
& \frac
    {\DS isBlocked(n,S,p) }
%   -----------------------------------------------------------
    {\DS \LL S, (p~and~break) \RR \ST \LL S, (clear(p)~;~break) \RR }
    & \textbf{(and-brk2)}   \\
%%%
& \LL S, (break~or~q) \ST \LL S, (clear(q)~;~break) \RR
    & \textbf{(or-brk1)}   \\
%%%
& \frac
    {\DS isBlocked(n,S,p) }
%   -----------------------------------------------------------
    {\DS \LL S, (p~or~break) \RR \ST \LL S, (clear(p)~;~break) \RR }
    & \textbf{(or-brk2)}   %\\
\end{eqnarray*}
}
%
A reaction eventually blocks in $awaiting$ and $emitting$ expressions in 
parallel trails.
%
If all trails hangs only in $awaiting$ expressions, it means that the program 
cannot advance in the current reaction.
%
However, $emitting$ expressions are pending in lower stack indexes and should 
eventualy resume in the ongoing reaction (see rule \textbf{pop}).
%
Therefore, we define another relation that behaves as \textbf{relation-inner} 
(presented above) if the program can advance, and, otherwise, pops the stack:
%
$$
\frac
    { \DS \LL S,p \RR \ST                   \LL S',p' \RR }
%   -----------------------------------------------------------
    {     \LL S,p \RR \xRightarrow[~~n~~]{} \LL S',p' \RR }
%
\5\5\5
%
\frac
    { \DS isBlocked(n,\1s:S,\1p) }
%   -----------------------------------------------------------
    { \LL s:S,p \RR \xRightarrow[~~n~~]{} \LL S,p \RR }
\5\5\5\textbf{(relation-outer)}
$$
%
To describe a \emph{reaction} in \CEU, i.e., how a program behaves in reaction 
to a single external event, we use the reflexive transitive closure of 
\textbf{relation-outer}:
%
$$
    \LL S,p \RR \xRightarrow[~~n~~]{*} \LL S',p' \RR
$$
%
Finally, to describe the complete execution of a program, we trigger multiple 
``invocations'' of reaction chains in sequence:
%
\begin{align*}
\LL [e1], p \RR
    & \xRightarrow[~~1~~]{*}
\LL [  ], p' \RR
\\
\LL [e2], p' \RR
    & \xRightarrow[~~2~~]{*}
\LL [  ], p'' \RR
\\
\LL [e3], p'' \RR
    & \xRightarrow[~~3~~]{*}
\LL [  ], p''' \RR
\\
& ...
\end{align*}
%
Each invocation starts with an external event at the top of the stack and 
finishes with a modified program and an empty stack.
After each invocation, we increment the sequence number.

@@

The concrete \code{await} and \code{emit} primitives support communication of 
values between them.
In the two-step translation of Figure~\ref{lst.map.emit.await}, we start with 
the concrete program in \CEU, which communicates the value $1$ between the 
\code{emit} and \code{await} in parallel (left-most code).
In the intermediate translation, we include the shared variable \code{e\_} to 
hold the value being communicated between the two trails in order to simplify 
the \code{emit}.
Finally, we convert the program into the equivalent in the abstract syntax, 
translating side-effect statements into $mem$ expressions.
External events require a similar translation, i.e., each external event has a 
corresponding variable that is explicitly set by the environment before each 
reaction.

@@

In the final generated code in \C, each trail segment label representing an 
entry point becomes a \emph{switch case} with the associated code to execute.
%
Figure~\ref{lst.impl.trails} illustrates the generation process.
For the program in the left of the figure, the compiler extracts the entry 
points and associated trails, e.g., the label \code{Awake\_e} will execute on 
\code{TRAIL-0} (ln. 7).
%
For each statement that pauses (\code{emit} and \code{await}), resumes 
(\code{par/and}, \code{par/or}, and \code{finalize}), or aborts (\code{par/or} 
and \code{break}), the compiler splits the trail into segments with associated 
entry points.
%
The entry points translate to an \code{enum} in the generated code (ln. 1--10, 
in the middle of the figure).
The state of trails translates to a vector of type \code{trail\_t} with the 
maximum number of simultaneous trails (ln. 12--15).
On initialization, \code{TRAIL-0} is set to execute the \code{Main} entry point 
(ln. 13), while all others are set to \code{INACTIVE} (in the example, only 
one, in ln. 14).

@@

The scheduler executes in two passes:
%
in the \emph{broadcast} pass, it sets all trails that are waiting for the 
current event to \code{STACKED} in the current stack level;
%
in the \emph{dispatch} pass, it executes each trail that is \code{STACKED} to 
run in the current level, setting it immediately to \code{INACTIVE} (the trail 
segment may reset it in sequence if it doesn't terminate).

During the dispatch pass, if a trail executes and emits an internal event, the 
scheduler increments the stack level and re-executes the two passes.
After all trails are properly dispatched, the scheduler decrements the stack 
level and resumes the previous execution.
For the first reaction, the scheduler starts from the \emph{dispatch} pass, 
given that the \code{Main} label is the only one that can be active at the 
stack level 0 (ln. 13, in the middle of Figure~\ref{lst.impl.trails}).

The code in the right of the Figure~\ref{lst.impl.trails} dispatches a trail 
segment according to the current label to execute.
%
For the first reaction, it executes the \code{Main} label in \code{TRAIL-0}.
%
When the \code{Main} label reaches the \code{par/and}, it stacks \code{TRAIL-1} 
(ln. 4--7) and proceeds to the code in \code{TRAIL-0} (ln. 9--14), respecting 
the deterministic execution order.
%
The code sets the running \code{TRAIL-0} to await \code{EVT\_e} on label 
\code{Awake\_e}, and then halts with a \code{break}.
%
The next iteration of \code{dispatch} takes \code{TRAIL-1} and executes its 
registered label \code{And\_sub\_2} (ln. 16--21), which sets \code{TRAIL-1} to 
await \code{EVT\_A} and also halts.

@@

\begin{description}
\item[{\textbf\code{ceu\_go\_init()}}:] initializes the program (e.g., the 
initial state of trails) and executes the ``boot'' reaction (i.e., the 
\code{Main} label).

\item[{\textbf\code{ceu\_go\_event(id,param)}}:] executes the reaction for the 
received external event id and associated parameter.

@@

The bindings for the specific platforms are responsible for calling the 
functions in the API in the order that better suit their requirements.
As an example, it is possible to set different priorities for events that occur 
concurrently (i.e. while a reaction is running).
However, a binding must never interleave or run multiple API calls in parallel.
This would break the \CEU sequential/discrete semantics of time.
%, as discussed in Section~\ref{sec:ceu}.

\begin{figure}
\begin{lstlisting}[numbers=left,xleftmargin=2em]
implementation
{
    #include "ceu_app.h"
    #include "ceu_app.c"

@@

As an example, Figure~\ref{lst.impl.tinyos} shows our binding for 
\emph{TinyOS}~\cite{wsn.tos}, which maps callbacks to input events in \CEU.
%
The file \code{ceu\_app.h} (ln. 3) contains all definitions for the compiled 
\CEU program, which are further queried through \code{\#ifdef}'s.
The file \code{ceu\_app.c} (ln. 4) contains the runtime of \CEU with the 
scheduler and dispatcher pointing to the labels defined in the program.
The callback \code{Boot.booted} (ln. 6--11) is called by TinyOS on startup, so 
we initialize \CEU inside it (ln. 7).
If the \CEU program uses timers, we also start a periodic timer (ln. 8--10) 
that triggers callback \code{Timer.fired} (ln. 13--17) every 10 milliseconds 
and advances the wall-clock time of \CEU (ln. 15)%
\footnote{We also offer a mechanism to start the underlying timer on demand to
avoid the ``battery unfriendly'' 10ms polling.}.
The remaining lines map pre-defined TinyOS events that can be used in \CEU 
programs, such as the light sensor (ln. 19--23) and the radio transceiver (ln.  
25--36).
%
The scheduler of the TinyOS is already synchronous by default and always 
execute \code{event} handlers atomically, hence, the API calls to \CEU are 
properly serialized.

@@

During compilation process the script is broken in trails in the same way it is 
done in the \CEU compiler explained in Section 4.
The Terra virtual machine implements a similar \CEU scheduler to execute the 
script trails, but the VM
uses the real trails address instead of labels. In this way is possible to jump directly to each script
partitions without using a dispatcher.
The Terra VM instruction set includes operations to support the \CEU execution 
model.
Basically, it is composed by instructions to define a pending trail (\code{tkins}),
a pending event(using \code{set\_c} to set a value in the memory),
a check operation to verify the \code{par/and} termination(\code{chkret}),
a pending timer event (\code{clken}), and some others like to clear trails data or suspend execution.
All pending entries are managed by the scheduler, that also responds for timer and input events.
All trails are executed as a TinyOS task, this ensure that each trail runs to completion avoiding race condition between then.

Figure~\ref{lst.impl.assembly} presents the bytecode generated for the initial 
part for the example of Figure~\ref{lst.impl.trails}.
After initializing the trail and event slots in the memory, the VM engine starts the execution from the \code{main} label.
In this example it registers in the trails slots the label addresses for the two block of the \code{par/and} control (lines 3--6).
The scheduler, first, executes the code in the label \code{and\_sub\_1} that registers the label \code{awk\_e} into the slot of event \code{e} (lines 8--10).
Then, executes the code in the label \code{and\_sub\_2} that registers the label \code{awk\_temp} into the slot of the event \code{I\_TEMP} (lines 17--18).
In this moment the VM engine suspends its execution waiting for some new input or timer event.
In the case of this example, the engine waits for the \code{I\_TEMP} event to continue the script execution.

\begin{figure}
\begin{minipage}[t]{0.91\linewidth}
\begin{lstlisting}[language=C,numbers=left,xleftmargin=2.5em,keywords={tkins,end,set_c,ushort,ubyte,exec}]
  <...> // Initializes the trails and events slots

:main -- Activate TRAILs 0 & 1
  tkins and_sub_1           // Activate TRAIL 0 - lbl and_sub_1
  tkins and_sub_2           // Activate TRAIL 1 - lbl and_sub_2
  end                       // Suspend execution

:and_sub_1
  set_c ushort slot_e awk_e // await e
  end                       // Suspend execution

:awk_e
  set_c ubyte slot_1 1      // Close TRAIL 0 - lbl and_sub_1
  exec chkret               // goto check par/and termination

:and_sub_2
  set_c ushort I_TEMP awk_temp // await TEMP
  end                       // Suspend execution

  <....> // Other entries: chkret, awk_TEMP, etc..
\end{lstlisting}
\end{minipage}
\caption{
%{\small %\textmd{
Assembly code for a \code{par/and} control structure.
%}%}
\label{lst.impl.assembly}
}
\end{figure}

@@

\begin{comment}
Instead of generating code in C, Terra generates the vmt assembly code using 
the original \CEU labels for trails entry points.
In the final phase for code generation, Terra compiler generates the vmt 
bytecode file using the program address in place of \CEU labels.
The use of address allows the \CEU scheduler to execute directly a trail 
without a index table to replace the switch-case.
Related to memory layout and trail allocation, Terra uses the same data 
organization approach for all \CEU controls and variables.
All \CEU data and control structures for a specific program are determined at 
compile time.
Terra arranges all these structures in the beginning of virtual machine memory 
space.
The original \CEU run time control functions are maintained with a small 
modification where accesses to control structures have an off-set relative to 
the begin of the virtual machine memory space.
\end{comment}

@@

\CEU has a strong influence from Esterel and embraces the disciplined 
synchronous-reactive model with support for lexical composition of lines of 
execution.
%
However, there are fundamental semantic differences that prevents the design of 
\CEU as pure extensions to Esterel.
%
In particular, Esterel has a notion of time similar to digital circuits in 
which multiple signals can be active at a clock tick.
In fact, Esterel is also used in hardware design.
%
In \CEU, instead of clock ticks, the occurrence of a single external event that 
defines a time unit.
%
\CEU also distinguishes external events from stack-based internal events, which 
provide a limited form of coroutines supporting reactive statements (e.g., 
\code{await} and \code{par/or}).

@@

A number of synchronous languages have been designed to interoperate with C, 
such as \emph{Reactive C}~\cite{rp.rc}, Protothreads~\cite{wsn.protothreads}, 
\emph{PRET-C}~\cite{rp.pretc} and \emph{SC}~\cite{rp.synchc}.
%
They offer Esterel-like parallel compositions with communication via shared 
variables, relying on deterministic scheduling to preserve determinism.
%
However, it is the responsibility of the programmer to specify the execution 
order for threads, based on either explicit priorities, or source code lexical 
order.
%
These languages have a tick-based notion of time similar to Esterel, which 
prevents the event-based temporal analysis of \CEU.

@@

Plummer et al.~\cite{esterel.vm} present a reactive virtual machine for 
embedded systems.
This VM uses the Esterel programming language as scripting language and implements
a specific instruction set to the Esterel concurrent and multi-thread semantic.
They focus in code reduction instead of on efficiency.
Experimentally, they found the bytecode was about half the size of optimized
native assembly code generated from C.
The Terra virtual machine also implements a customized instruction set for a reactive
programming model, but it uses the \CEU model for concurrence and event 
control.
The Terra component based approach also contributes to have a very small bytecode size.
In our experiments, depending of the abstraction level of the pre-built customized components,
the bytecode size reduction may reach about 99\%.

@@

\CEU is a concurrency-safe language, employing a static analysis that encompass 
all control constructs and ensures that the high degree of concurrency in 
embedded systems does not pose safety threats to applications.
%
As a summary, the following safety properties hold for all programs that 
successfully compile in \CEU:
time and memory-bounded reactions to the environment (except for external 
system calls),
no race conditions in shared memory,
reliable abortion for activities handling resources,
and automatic synchronization for timers.
%
These properties are usually desirable in embedded applications and are 
guaranteed as preconditions in \CEU by design.
